---
title: "ZmFd4 / ZmFd9 promoter primers across PanAnd teosintes"
subtitle: "Orthologs → loci → alignment → conserved primer sites → Primer3 (2.5–3 kb) → per-accession binding + whole-genome e-PCR"
author: "thegemmalab@ncsu.edu"
date: today
params:
  mode: stub          # stub = laptop fixtures (B73 + TIL18 slices); full = all genomes on hazel
format:
  html:
    output-file: fd_promoter_primers.html
    toc: true
    toc-depth: 2
    code-fold: true
    embed-resources: true
    fig-width: 12
    fig-height: 5
execute:
  warning: false
  message: false
  echo: true
---

**Goal.** Primers that amplify the region upstream of the ATG of the maize ferredoxins **ZmFd4**
(`Zm00001eb083950`) and **ZmFd9** (`Zm00001eb421930`) in every published PanAnd *Zea* teosinte
assembly, so the region can be resequenced (cf. Jia *et al.* 2025, who resequenced a 2.9-kb ZmFd4
fragment and found InDel194, a 604-bp deletion upstream of the ATG).

**Design rules.**

- **Reverse primer** in the **first CDS exon** (in B73 v5 both genes start with a non-coding 5′UTR
  exon; the ATG is in exon 2, which holds the whole CDS).
- Primer sites are chosen by **conservation, not coordinates**: a primer must sit on alignment
  columns that are identical and gap-free in **every** accession (pass 1). Only if pass 1 finds
  nothing are SNP columns allowed, as IUPAC degenerate bases, with an invariant 3′ 5 nt (pass 2).
- **Product 2,500–3,000 bp** (Primer3), forward primer upstream of the ATG.
- **Specificity:** forward e-PCR over each whole genome, as in `primermu`.

```{r setup}
suppressPackageStartupMessages({
  library(Biostrings)                       # first, so dplyr/purrr verbs (slice, filter, reduce) win
  library(gggenomes); library(ggplot2); library(readr); library(dplyr); library(tidyr)
  library(stringr); library(purrr)
})

MODE <- params$mode
stopifnot(MODE %in% c("stub", "full"))
REF  <- if (MODE == "stub") "tests/fixtures" else "/rsstu/users/r/rrellan/BZea/ref"
ACC  <- read_tsv(if (MODE == "stub") "config/accessions_stub.tsv" else "config/accessions.tsv",
                 show_col_types = FALSE)
WORK <- file.path("work", MODE)
RES  <- if (MODE == "stub") file.path(WORK, "results") else "results"
for (d in file.path(WORK, c("", "fai", "blastdb", "genomes", "loci", "aln", "p3", "epcr")))
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
dir.create(RES, showWarnings = FALSE, recursive = TRUE)
THREADS <- as.integer(Sys.getenv("SLURM_CPUS_PER_TASK", "4"))

GENES <- tibble(gene = c("ZmFd4", "ZmFd9"), b73_id = c("Zm00001eb083950", "Zm00001eb421930"))
UP <- 6000L; DOWN <- 1000L                 # locus = ATG - UP .. CDS end + DOWN, in gene orientation
PRODUCT <- c(2500L, 3000L)

sh  <- function(cmd) system2("bash", c("-c", shQuote(cmd)), stdout = TRUE, stderr = TRUE)
out <- function(cmd) system2("bash", c("-c", shQuote(cmd)), stdout = TRUE, stderr = FALSE)
tools <- c("samtools", "blastn", "makeblastdb", "mafft", "primer3_core", "e-PCR")
miss <- tools[Sys.which(tools) == ""]
if (length(miss)) stop("missing tools: ", paste(miss, collapse = ", "))
cat(sprintf("mode = %s | %d accessions | reference dir %s\n", MODE, nrow(ACC), REF))
```

## 1. Genomes: plain FASTA, index, BLAST database

Genomes are read in place from `r REF`; only a gzipped genome (RIL003) is decompressed into
`work/` (e-PCR needs plain FASTA). FASTA indexes and any missing BLAST DB are written to `work/`,
never next to the shared references.

```{r genomes}
ACC <- ACC %>% mutate(
  fa_src = file.path(REF, genome),
  fa     = ifelse(grepl("\\.gz$", genome), file.path(WORK, "genomes", paste0(accession, ".fa")), fa_src),
  fai    = file.path(WORK, "fai", paste0(accession, ".fa.fai")),
  gff    = file.path(REF, gff),
  db_ref = sub("\\.fa(\\.gz)?$", "", fa_src))
stopifnot(all(file.exists(ACC$fa_src)))

for (i in seq_len(nrow(ACC))) with(ACC[i, ], {
  if (fa != fa_src && !file.exists(fa)) sh(sprintf("gzip -dc %s > %s", shQuote(fa_src), shQuote(fa)))
  if (!file.exists(fai)) sh(sprintf("samtools faidx --fai-idx %s %s", shQuote(fai), shQuote(fa)))
})
has_db <- function(p) any(file.exists(paste0(p, c(".nsq", ".nal", ".00.nsq"))))
ACC$db <- map2_chr(ACC$db_ref, ACC$accession, function(p, a) {
  if (has_db(p)) return(p)
  w <- file.path(WORK, "blastdb", a)
  if (!has_db(w)) sh(sprintf("makeblastdb -dbtype nucl -in %s -out %s",
                             shQuote(ACC$fa[ACC$accession == a]), shQuote(w)))
  w
})

grab <- function(acc, chr, s, e, rc = FALSE) {
  a <- ACC[ACC$accession == acc, ]
  o <- out(sprintf("samtools faidx --fai-idx %s %s %s:%d-%d", shQuote(a$fai), shQuote(a$fa), chr, s, e))
  d <- DNAString(toupper(paste(o[-1], collapse = "")))
  if (rc) reverseComplement(d) else d
}
gff_cols <- c("seqid","src","type","start","end","score","strand","phase","attr")
read_gff <- function(gff, filter_awk) {
  cat_cmd <- if (grepl("\\.gz$", gff)) "gzip -dc" else "cat"
  txt <- out(sprintf("%s %s | awk -F'\\t' '%s'", cat_cmd, shQuote(gff), filter_awk))
  if (!length(txt)) return(NULL)
  read_tsv(I(txt), col_names = gff_cols, col_types = "ccciicccc") %>%
    mutate(ID = str_match(attr, "(?:^|;)ID=([^;]+)")[, 2], Parent = str_match(attr, "Parent=([^;]+)")[, 2])
}
knitr::kable(ACC %>% transmute(accession, taxon, annotation, blast_db = basename(db)))
```

## 2. B73 query: the CDS of each gene

```{r b73-cds}
b73 <- ACC %>% filter(accession == "B73")
b73_models <- map_dfr(GENES$b73_id, function(id)
  read_gff(b73$gff, sprintf("$9 ~ /%s/", id)) %>% mutate(b73_id = id))
cds_q <- GENES %>% rowwise() %>% mutate(seq = {
  cds <- b73_models %>% filter(b73_id == .env$b73_id, type == "CDS", grepl("_T001", Parent))
  rc <- cds$strand[1] == "-"
  cds <- if (rc) arrange(cds, desc(start)) else arrange(cds, start)
  as.character(do.call(xscat, map(seq_len(nrow(cds)), ~ grab("B73", cds$seqid[.x], cds$start[.x], cds$end[.x], rc))))
}) %>% ungroup()
qfa <- file.path(WORK, "cds_queries.fa")
writeLines(paste0(">", cds_q$gene, "\n", cds_q$seq), qfa)
knitr::kable(cds_q %>% transmute(gene, b73_id, cds_bp = nchar(seq),
                                 start = substr(seq, 1, 3), end = substring(seq, nchar(seq) - 2)))
```

## 3. Orthologs: one locus per gene per accession

`blastn` of both B73 CDSs against every genome. For each gene the ortholog is the best hit covering
≥ 90 % of the CDS. **Paralog guard:** that locus must score higher with its own CDS than with the
paralog's (Fd4 and Fd9 CDSs are ~85 % identical, so a teosinte Fd9 is also a decent Fd4 hit). The
PanAnd gene model overlapping the hit gives the ortholog's ID and exon structure.

```{r orthologs}
bl_cols <- c("qseqid","sseqid","pident","length","qstart","qend","sstart","send","evalue","bitscore","qlen")
hits <- map_dfr(seq_len(nrow(ACC)), function(i) {
  o <- out(sprintf("blastn -task megablast -query %s -db %s -evalue 1e-20 -max_target_seqs 20 -num_threads %d -outfmt '6 %s'",
                   shQuote(qfa), shQuote(ACC$db[i]), THREADS, paste(bl_cols, collapse = " ")))
  if (!length(o)) return(NULL)
  read_tsv(I(o), col_names = bl_cols, show_col_types = FALSE) %>% mutate(accession = ACC$accession[i])
}) %>% mutate(strand = ifelse(send >= sstart, "+", "-"), lo = pmin(sstart, send), hi = pmax(sstart, send),
              qcov = length / qlen)

orth <- map_dfr(seq_len(nrow(ACC)), function(i) map_dfr(GENES$gene, function(g) {
  a <- ACC$accession[i]
  cand <- hits %>% filter(accession == a, qseqid == g, qcov >= 0.9) %>% arrange(desc(bitscore))
  if (!nrow(cand)) return(tibble(accession = a, gene = g, status = "no hit"))
  b <- cand[1, ]
  para <- hits %>% filter(accession == a, qseqid != g, sseqid == b$sseqid, hi >= b$lo, lo <= b$hi)
  para_bits <- if (nrow(para)) max(para$bitscore) else 0
  atg <- if (b$strand == "+") b$sstart - (b$qstart - 1L) else b$sstart + (b$qstart - 1L)
  cend <- if (b$strand == "+") b$send + (b$qlen - b$qend) else b$send - (b$qlen - b$qend)
  tibble(accession = a, gene = g, chr = b$sseqid, strand = b$strand, atg = atg, cds_end = cend,
         pident = b$pident, bits = b$bitscore, paralog_bits = para_bits,
         next_bits = if (nrow(cand) > 1) cand$bitscore[2] else NA_real_,
         status = ifelse(b$bitscore > para_bits, "ok", "PARALOG?"))
}))

# PanAnd / Helixer gene model overlapping each hit
models <- map_dfr(seq_len(nrow(orth)), function(k) {
  o <- orth[k, ]; if (o$status == "no hit") return(NULL)
  g <- ACC$gff[ACC$accession == o$accession]
  lo <- min(o$atg, o$cds_end); hi <- max(o$atg, o$cds_end)
  gm <- read_gff(g, sprintf("$1==\"%s\" && $5>=%d && $4<=%d", o$chr, lo - 20000L, hi + 20000L))
  if (is.null(gm)) return(NULL)
  gmod <- gm %>% filter(type == "gene", end >= lo, start <= hi, strand == o$strand) %>% slice(1)
  if (!nrow(gmod)) return(NULL)
  tx <- gm %>% filter(type == "mRNA", Parent == gmod$ID) %>%
    arrange(desc(grepl("canonical_transcript=1", attr))) %>% slice(1)
  gm %>% filter(Parent == tx$ID, type %in% c("exon", "CDS", "five_prime_UTR", "three_prime_UTR")) %>%
    mutate(accession = o$accession, gene = o$gene, model_id = gmod$ID)
})
orth <- orth %>% left_join(models %>% distinct(accession, gene, model_id), by = c("accession", "gene"))
write_tsv(orth, file.path(RES, "orthologs.tsv"))
knitr::kable(orth %>% transmute(gene, accession, model_id, locus = sprintf("%s:%s (%s)", chr, format(atg, big.mark = ","), strand),
                                pident, bits, paralog_bits, status),
             caption = "ATG position of each ortholog (from the CDS hit). paralog_bits = best score of the other gene's CDS at the same locus.")
stopifnot(all(orth$status == "ok"))
```

## 4. Loci in gene orientation

Each locus runs from `r UP` bp upstream of the ATG to `r DOWN` bp past the CDS end, reverse
complemented for minus-strand genes, so position `r UP + 1` is the A of the ATG in every accession.
Gene-model features are projected onto the same coordinates.

```{r loci}
orth <- orth %>% mutate(
  from = ifelse(strand == "+", atg - UP, cds_end - DOWN),
  to   = ifelse(strand == "+", cds_end + DOWN, atg + UP))
to_locus <- function(g, o) if (o$strand == "+") g - o$from + 1L else o$to - g + 1L

loci <- map(set_names(GENES$gene), function(g) {
  oo <- orth %>% filter(gene == g)
  s <- DNAStringSet(map(seq_len(nrow(oo)), function(k) grab(oo$accession[k], oo$chr[k], oo$from[k], oo$to[k], oo$strand[k] == "-")))
  names(s) <- oo$accession
  writeXStringSet(s, file.path(WORK, "loci", paste0(g, ".fa")))
  s
})
feats <- models %>% left_join(orth %>% select(accession, gene, strand, from, to), by = c("accession", "gene")) %>%
  rowwise() %>% mutate(s = to_locus(start, list(strand = strand.y, from = from, to = to)),
                       e = to_locus(end,   list(strand = strand.y, from = from, to = to))) %>% ungroup() %>%
  transmute(gene, accession, type, start = pmin(s, e), end = pmax(s, e)) %>% arrange(gene, accession, start)
# first CDS exon of each accession (transcript order = locus order after orientation)
cds1 <- feats %>% filter(type == "CDS") %>% group_by(gene, accession) %>% slice_min(start, n = 1) %>% ungroup()
knitr::kable(feats %>% filter(type == "exon") %>% group_by(gene, accession) %>%
               summarise(exons = paste(sprintf("%d-%d", start - UP - 1L, end - UP - 1L), collapse = ", "), .groups = "drop"),
             caption = "Exons relative to the ATG (A = 0). The 5′UTR exon and intron lie upstream of the ATG.")
```

## 5. Alignment and conserved columns

MAFFT E-INS-i (`--genafpair`, suited to long unalignable stretches such as indels/TEs) per gene.
Conservation is evaluated on **B73 coordinates** of the alignment:

- **invariant** = every accession has the B73 base, no gap, and no insertion in any accession
  between this base and the next (an insertion would break a primer's contiguous match);
- **SNP** = all accessions have a base (no gap/insertion) but not the same one — allowed only in pass 2.

```{r align}
aln <- map(set_names(GENES$gene), function(g) {
  af <- file.path(WORK, "aln", paste0(g, ".aln.fa"))
  if (!file.exists(af)) sh(sprintf("mafft --genafpair --maxiterate 1000 --thread %d --quiet %s > %s",
                                   THREADS, shQuote(file.path(WORK, "loci", paste0(g, ".fa"))), shQuote(af)))
  m <- as.matrix(readDNAMultipleAlignment(af))
  m[c("B73", setdiff(rownames(m), "B73")), , drop = FALSE]
})

cons <- imap(aln, function(m, g) {
  cols <- which(m["B73", ] != "-"); n <- length(cols)
  col_ok <- function(cc) { x <- m[, cc]; c(all(x != "-" & x != "N"), length(unique(x)) == 1L) }
  st <- vapply(cols, col_ok, logical(2))
  ins_after <- vapply(seq_len(n), function(i) {
    if (i == n) return(FALSE); between <- seq_len(cols[i + 1] - cols[i] - 1) + cols[i]
    length(between) > 0 && any(m[, between, drop = FALSE] != "-") }, logical(1))
  brk <- ins_after | c(FALSE, ins_after[-n])
  # per-column entropy over bases only (gaps tracked separately so a large indel is not "entropy")
  ent <- vapply(cols, function(cc) { x <- m[, cc]; x <- x[x %in% c("A","C","G","T")]
    if (!length(x)) return(NA_real_); p <- table(x) / length(x); -sum(p * log2(p)) }, numeric(1))
  gapf <- vapply(cols, function(cc) mean(m[, cc] == "-"), numeric(1))
  tibble(gene = g, pos = seq_len(n), b73 = m["B73", cols],
         status = case_when(!st[1, ] | brk ~ "gap", st[2, ] ~ "invariant", TRUE ~ "snp"),
         entropy = ent, gap_frac = gapf, col = cols)
})
cons_all <- bind_rows(cons)
knitr::kable(cons_all %>% count(gene, status) %>% pivot_wider(names_from = status, values_from = n),
             caption = "B73 positions of each locus by cross-accession status.")
```

```{r cons-plot, fig.height=4}
cons_all %>% mutate(x = pos - UP - 1L) %>% group_by(gene) %>%
  mutate(invariant = stats::filter(status == "invariant", rep(1/50, 50), sides = 2),
         gaps = stats::filter(status == "gap", rep(1/50, 50), sides = 2)) %>%
  select(gene, x, invariant, gaps) %>% pivot_longer(c(invariant, gaps)) %>%
  ggplot(aes(x, value, colour = name)) + geom_line(linewidth = 0.4) + facet_wrap(~gene, ncol = 1) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "position relative to ATG (bp, B73)", y = "fraction in 50-bp window", colour = NULL,
       title = "Conservation along each locus") + theme_bw()
```

## 6. Primer3 on the conserved template

The B73 locus is the template, with every non-invariant position replaced by `N` and
`PRIMER_MAX_NS_ACCEPTED=0`, so Primer3 can only place primers on sites identical in all
accessions. `SEQUENCE_PRIMER_PAIR_OK_REGION_LIST` confines the forward primer to the region
upstream of the ATG and the reverse primer to the first CDS exon.

```{r p3}
iupac <- c(A="A",C="C",G="G",T="T",AG="R",CT="Y",CG="S",AT="W",GT="K",AC="M",CGT="B",AGT="D",ACT="H",ACG="V",ACGT="N")
iupac_col <- function(x) iupac[[paste(sort(unique(x)), collapse = "")]]
fold_deg <- function(s) { k <- c(R=2,Y=2,S=2,W=2,K=2,M=2,B=3,D=3,H=3,V=3,N=4)[strsplit(s, "")[[1]]]; prod(k, na.rm = TRUE) }
deg_primer <- function(m, cols, rev) {
  s <- vapply(cols, function(cc) iupac_col(m[, cc]), "")
  s <- paste(s, collapse = ""); if (rev) as.character(reverseComplement(DNAString(s))) else s
}
run_p3 <- function(g, pass) {
  cs <- cons[[g]]; b <- cds1 %>% filter(gene == g, accession == "B73")
  keep <- if (pass == 1) cs$status == "invariant" else cs$status != "gap"
  tmpl <- paste(ifelse(keep, cs$b73, "N"), collapse = "")
  p3in <- file.path(WORK, "p3", sprintf("%s_pass%d.in", g, pass)); p3out <- sub("\\.in$", ".out", p3in)
  writeLines(c(sprintf("SEQUENCE_ID=%s_pass%d", g, pass), paste0("SEQUENCE_TEMPLATE=", tmpl),
    sprintf("SEQUENCE_PRIMER_PAIR_OK_REGION_LIST=0,%d,%d,%d", UP, b$start - 1L, b$end - b$start + 1L),
    "PRIMER_TASK=generic", "PRIMER_FIRST_BASE_INDEX=0", "PRIMER_PICK_INTERNAL_OLIGO=0",
    sprintf("PRIMER_PRODUCT_SIZE_RANGE=%d-%d", PRODUCT[1], PRODUCT[2]),
    "PRIMER_MIN_SIZE=19", "PRIMER_OPT_SIZE=22", "PRIMER_MAX_SIZE=27",
    "PRIMER_MIN_TM=57.0", "PRIMER_OPT_TM=60.0", "PRIMER_MAX_TM=63.0", "PRIMER_PAIR_MAX_DIFF_TM=3.0",
    "PRIMER_MIN_GC=40.0", "PRIMER_OPT_GC_PERCENT=50.0", "PRIMER_MAX_GC=60.0",
    "PRIMER_GC_CLAMP=1", "PRIMER_MAX_POLY_X=4", "PRIMER_MAX_NS_ACCEPTED=0",
    "PRIMER_EXPLAIN_FLAG=1", "PRIMER_NUM_RETURN=50", "="), p3in)
  sh(sprintf("primer3_core %s > %s", shQuote(p3in), shQuote(p3out)))
  L <- readLines(p3out); L <- L[grepl("=", L, fixed = TRUE) & L != "="]
  kv <- strsplit(L, "=", fixed = TRUE)
  d <- setNames(vapply(kv, function(x) paste(x[-1], collapse = "="), ""), vapply(kv, `[`, "", 1))
  np <- as.integer(d[["PRIMER_PAIR_NUM_RETURNED"]])
  if (np == 0) { message(g, " pass ", pass, ": ", d[["PRIMER_PAIR_EXPLAIN"]]); return(NULL) }
  m <- aln[[g]]
  map_dfr(0:(np - 1), function(i) {
    v <- function(k) d[[sprintf(k, i)]]
    lp <- as.integer(strsplit(v("PRIMER_LEFT_%d"), ",")[[1]]); rp <- as.integer(strsplit(v("PRIMER_RIGHT_%d"), ",")[[1]])
    f <- (lp[1] + 1L):(lp[1] + lp[2]); r <- (rp[1] - rp[2] + 2L):(rp[1] + 1L)     # 1-based B73 locus positions
    Fd <- deg_primer(m, cs$col[f], FALSE); Rd <- deg_primer(m, cs$col[r], TRUE)
    tibble(gene = g, pass = pass, rank = i, penalty = round(as.numeric(v("PRIMER_PAIR_%d_PENALTY")), 3),
      prod_b73 = as.integer(v("PRIMER_PAIR_%d_PRODUCT_SIZE")),
      F_seq = Fd, R_seq = Rd, F_len = lp[2], R_len = rp[2],
      F_tm = round(as.numeric(v("PRIMER_LEFT_%d_TM")), 1), R_tm = round(as.numeric(v("PRIMER_RIGHT_%d_TM")), 1),
      F_gc = round(as.numeric(v("PRIMER_LEFT_%d_GC_PERCENT")), 1), R_gc = round(as.numeric(v("PRIMER_RIGHT_%d_GC_PERCENT")), 1),
      F_from_atg = min(f) - UP - 1L, R_from_atg = max(r) - UP - 1L,
      deg = str_count(paste0(Fd, Rd), "[^ACGT]"),
      fold = fold_deg(paste0(Fd, Rd)),
      end3_ok = !grepl("[^ACGT]", str_sub(Fd, -5)) & !grepl("[^ACGT]", str_sub(Rd, -5)))
  })
}
pairs <- map_dfr(GENES$gene, function(g) {
  p <- run_p3(g, 1)
  if (is.null(p)) p <- run_p3(g, 2) %>% filter(end3_ok, deg <= 4, fold <= 16)
  p
})
stopifnot(nrow(pairs) > 0)
```

## 7. Every primer must bind every accession

Each primer (degenerate primers as IUPAC patterns) is matched exactly against each accession's
locus; the product size per accession follows from the two sites.

```{r binding}
bind_check <- pairs %>% rowwise() %>% mutate(chk = list(map_dfr(names(loci[[gene]]), function(a) {
  s <- loci[[gene]][[a]]
  fh <- start(matchPattern(DNAString(F_seq), s, fixed = "subject"))
  rh <- end(matchPattern(reverseComplement(DNAString(R_seq)), s, fixed = "subject"))
  tibble(accession = a, F_sites = length(fh), R_sites = length(rh),
         product = if (length(fh) == 1 && length(rh) == 1) rh - fh + 1L else NA_integer_)
}))) %>% ungroup() %>%
  mutate(all_bind = map_lgl(chk, ~ all(.x$F_sites == 1 & .x$R_sites == 1)),
         min_prod = map_int(chk, ~ suppressWarnings(min(.x$product))),
         max_prod = map_int(chk, ~ suppressWarnings(max(.x$product))))
cand <- bind_check %>% filter(all_bind) %>% group_by(gene) %>% arrange(penalty, .by_group = TRUE) %>%
  slice_head(n = 5) %>% mutate(name = sprintf("%s-P%d", gene, row_number())) %>% ungroup()
knitr::kable(bind_check %>% count(gene, all_bind), caption = "Primer3 pairs whose both primers bind once in every accession.")
```

## 8. Whole-genome e-PCR specificity

`-n 3 -g 1`: up to 3 mismatches and 1 gap per primer (as in `primermu`). A good pair gives
**one** product per genome, at its own locus, and nothing at the paralog.

```{r epcr}
jia <- read_tsv("config/jia2025_primers.tsv", show_col_types = FALSE)
jia_pairs <- tibble(name = c("Jia-G1 (G-F1/G-R1)", "Jia-G2 (G-F2/G-R2)", "Jia-Fd9-JD"), gene = c("ZmFd4", "ZmFd4", "ZmFd9"),
                    F_seq = jia$seq[match(c("ZmFd4-G-F1", "ZmFd4-G-F2", "ZmFd9-JD-F"), jia$name)],
                    R_seq = jia$seq[match(c("ZmFd4-G-R1", "ZmFd4-G-R2", "ZmFd9-JD-R"), jia$name)])
sts_tbl <- bind_rows(cand %>% select(name, gene, F_seq, R_seq), jia_pairs)
sts <- file.path(WORK, "epcr", "pairs.sts")
writeLines(sprintf("%s\t%s\t%s\t100-5000", str_replace_all(sts_tbl$name, "[ ()/]", "_"), sts_tbl$F_seq, sts_tbl$R_seq), sts)
ep <- map_dfr(seq_len(nrow(ACC)), function(i) {
  o <- file.path(WORK, "epcr", paste0(ACC$accession[i], ".txt"))
  if (!file.exists(o)) sh(sprintf("e-PCR -n 3 -g 1 -m 50 -w 7 -t 3 -a f -o %s %s %s", shQuote(o), shQuote(sts), shQuote(ACC$fa[i])))
  if (!file.info(o)$size) return(NULL)
  read_tsv(o, comment = "#", show_col_types = FALSE, col_names = c("seq","sts","strand","from","to","size","mism","gaps","x")) %>%
    mutate(accession = ACC$accession[i])
}) %>% left_join(sts_tbl %>% mutate(sts = str_replace_all(name, "[ ()/]", "_")) %>% select(sts, name, pair_gene = gene), by = "sts")
# where is each product: own locus, paralog locus, elsewhere
ep <- ep %>% rowwise() %>% mutate(hit_at = {
  o <- orth %>% filter(accession == .env$accession, chr == seq, from <= .env$to, to >= .env$from)
  if (!nrow(o)) "elsewhere" else if (o$gene[1] == pair_gene) "own locus" else "PARALOG"
}) %>% ungroup()
spec <- ep %>% count(name, accession, hit_at) %>% pivot_wider(names_from = hit_at, values_from = n, values_fill = 0)
write_tsv(ep, file.path(RES, "epcr_hits.tsv"))
knitr::kable(spec, caption = "e-PCR products per pair and genome.")
```

## 9. Recommended pairs

```{r final}
specific <- spec %>% group_by(name) %>%
  summarise(single = all(`own locus` == 1) & all(if ("PARALOG" %in% names(spec)) PARALOG == 0 else TRUE) &
                     all(if ("elsewhere" %in% names(spec)) elsewhere == 0 else TRUE), .groups = "drop")
final <- cand %>% left_join(specific, by = "name") %>% filter(single %in% TRUE)
write_tsv(final %>% select(name, gene, pass, F_seq, R_seq, F_len, R_len, F_tm, R_tm, F_gc, R_gc,
                           F_from_atg, R_from_atg, prod_b73, min_prod, max_prod, deg, penalty),
          file.path(RES, "primers.tsv"))
knitr::kable(final %>% transmute(name, forward = F_seq, reverse = R_seq, `F Tm` = F_tm, `R Tm` = R_tm,
                                 `F 5′ vs ATG` = F_from_atg, `R 3′ end vs ATG` = R_from_atg,
                                 `B73 bp` = prod_b73, `range across accessions` = sprintf("%d–%d", min_prod, max_prod),
                                 degenerate = deg, penalty),
             caption = "Pairs that bind every accession once and give a single whole-genome e-PCR product in every genome.")
```

## 10. Jia *et al.* 2025 primers as a benchmark

Jia's primers mapped onto every locus (up to 3 mismatches). For ZmFd4, the G-F1/G-R1 product size
calls the InDel194 allele group: full promoter (Hap1) vs. the 604-bp deletion (Hap2).

```{r jia}
jia_map <- jia %>% filter(gene %in% GENES$gene) %>% rowwise() %>% mutate(m = list(map_dfr(names(loci[[gene]]), function(a) {
  s <- loci[[gene]][[a]]; p <- DNAString(seq)
  hf <- matchPattern(p, s, max.mismatch = 3); hr <- matchPattern(reverseComplement(p), s, max.mismatch = 3)
  h <- c(start(hf), end(hr)); mm <- c(nmismatch(p, hf), nmismatch(reverseComplement(p), hr))
  tibble(accession = a, sites = length(h), pos = if (length(h)) h[which.min(mm)] - UP - 1L else NA_integer_,
         mism = if (length(h)) min(mm) else NA_integer_)
}))) %>% ungroup() %>% unnest(m)
knitr::kable(jia_map %>% transmute(name, accession, mism = ifelse(is.na(mism), "no site (>3 mm)", as.character(mism))) %>%
               pivot_wider(names_from = accession, values_from = mism),
             caption = "Mismatches of each Jia primer at its best site in each locus.")
g1 <- jia_map %>% filter(name %in% c("ZmFd4-G-F1", "ZmFd4-G-R1")) %>% select(name, accession, pos) %>%
  pivot_wider(names_from = name, values_from = pos) %>% mutate(G1_bp = `ZmFd4-G-R1` - `ZmFd4-G-F1` + 1L)
knitr::kable(g1 %>% transmute(accession, `G-F1/G-R1 product (bp)` = G1_bp,
                              InDel194 = case_when(is.na(G1_bp) ~ "?", G1_bp > 1700 ~ "Hap1 (full)", TRUE ~ "Hap2 (deletion)")),
             caption = "ZmFd4 InDel194 call from Jia's G-F1/G-R1 product (B73 Hap1 ≈ 1,908 bp; Hap2 ≈ 604 bp shorter).")
```

## 11. Maps

```{r maps, fig.height=7}
draw_gene <- function(g) {
  ls <- loci[[g]]
  seqs <- tibble(seq_id = names(ls), length = Biostrings::width(ls))
  ft <- feats %>% filter(gene == g, type %in% c("exon", "CDS")) %>%
    transmute(seq_id = accession, start, end, feat_type = type)
  fp <- final %>% filter(gene == g) %>% slice_head(n = 1)
  prim <- map_dfr(names(ls), function(a) {
    s <- ls[[a]]
    if (!nrow(fp)) return(NULL)
    f <- matchPattern(DNAString(fp$F_seq), s, fixed = "subject"); r <- matchPattern(reverseComplement(DNAString(fp$R_seq)), s, fixed = "subject")
    tibble(seq_id = a, start = c(start(f), start(r)), end = c(end(f), end(r)), feat_type = c("forward primer", "reverse primer"))
  })
  # alignment blocks between adjacent tracks from pairwise blastn
  links <- map_dfr(seq_len(length(ls) - 1), function(k) {
    a <- tempfile(fileext = ".fa"); b <- tempfile(fileext = ".fa")
    writeXStringSet(ls[k], a); writeXStringSet(ls[k + 1], b)
    o <- out(sprintf("blastn -query %s -subject %s -evalue 1e-20 -outfmt '6 qstart qend sstart send pident'", a, b))
    if (!length(o)) return(NULL)
    read_tsv(I(o), col_names = c("start","end","start2","end2","pident"), show_col_types = FALSE) %>%
      filter(start2 < end2) %>% mutate(seq_id = names(ls)[k], seq_id2 = names(ls)[k + 1])
  })
  gggenomes(seqs = seqs, feats = list(ft, prim), links = links) +
    geom_link(aes(fill = pident), alpha = 0.35, colour = NA) +
    geom_seq() + geom_bin_label() +
    geom_feat(data = feats(ft), aes(colour = feat_type), linewidth = 4, position = "identity") +
    geom_feat(data = feats(prim), aes(colour = feat_type), linewidth = 6, position = "identity") +
    scale_colour_manual(values = c(exon = "#DAA520", CDS = "#8B6914", "forward primer" = "#1F6FB4", "reverse primer" = "#B4451F"), name = NULL) +
    scale_fill_viridis_c(name = "% identity", option = "C") +
    labs(title = sprintf("%s locus, ATG at %d (gene orientation)", g, UP + 1L),
         subtitle = if (nrow(fp)) sprintf("%s: %s / %s", fp$name, fp$F_seq, fp$R_seq) else "no pair passed") +
    theme(legend.position = "bottom")
}
for (g in GENES$gene) { p <- draw_gene(g); ggsave(file.path(RES, sprintf("map_%s.png", g)), p, width = 12, height = 7, dpi = 150); print(p) }
```

## Next / TODO (student)

- Motif scan of the amplified region (FIMO + MAST, JASPAR 2024 CORE Plantae); genome-specific
  background (`fasta-get-markov`); nitrogen motif set (NLP / NIGT1).
- Variant catalogue and sequence classes of the amplified region; LD among variants.
- Synteny confirmation of orthologs (neighbouring genes), beyond the paralog score guard.
- Population check with the MaizeGDB 2026 teosinte VCFs.

```{r sessioninfo}
#| code-fold: true
sessionInfo()
```
